Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene

### 2023-02-05 manuscript/manuscript.Rmd
update introduction

Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. nearest living relative...

We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.

@@

>>>>

Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SSTs) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. nearest living relative...

We apply this approach to the record of the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012]. The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. Although some more recent climate models have been able to align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate. A relatively recent, proxy-based studies inferred a SST gradient that was shallower than the modern by 32+-10 % [@Evans2018], which would correspond to a gradient of ~20+-3 degC. Later studies based on new proxy compilations only estimate temperature gradients up to mid-latitudes [@Hollis2019], or combine SST proxies with other data [@Inglis2020], making it difficult to infer a meaningful estimate of the SST gradient.

Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.

@@

-GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
